---
title: "Generate ALS cohort descriptive statistics and illustrative timeline figures"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load necessary libraries
library(tidyverse)
library(tableone)
library(MASS)
library(broom)
library(JMbayes)
library(splines)
library(knitr)
library(patchwork)

# Load data
df_ind <- read.csv("Data/casedata.csv", stringsAsFactors = FALSE)
df_long <- read.csv("Data/longdata.csv", stringsAsFactors = FALSE)


# Set theme for graphs
theme.apply <- theme(panel.background = element_rect(fill = 'white'),
                     panel.grid.major.y = element_line(colour = "grey90", size=0.5),
                     panel.grid.major.x = element_line(colour = "grey95", size=0.5),
                     plot.title = element_text(size=14, face="bold"),
                     legend.title = element_text(size=16, face="bold"),
                     axis.title.x = element_text(size=14),
                     axis.title.y = element_text(size=14, vjust=0.5),
                     axis.text.x = element_text(size=14, face="bold"),
                     axis.text.y = element_text(size=14, face="bold", angle=25))

theme.apply_noleg <- theme(panel.background = element_rect(fill = 'white'),
                     panel.grid.major.y = element_line(colour = "grey90", size=0.5),
                     panel.grid.major.x = element_line(colour = "grey95", size=0.5),
                     plot.title = element_text(size=14, face="bold"),
                     legend.position = "none",
                     axis.title.x = element_text(size=14),
                     axis.title.y = element_text(size=14, vjust=0.5),
                     axis.text.x = element_text(size=14, face="bold"),
                     axis.text.y = element_text(size=14, face="bold", angle=25))


##############################

# 1. Convert strings to factors
df_ind$simp_site <- factor(df_ind$simp_site, levels = c("Spinal", "Bulbar"))
df_long$simp_site <- factor(df_long$simp_site, levels = c("Spinal", "Bulbar"))
df_long$ID <- as.factor(df_long$ID)

# 2. Add small value to survival time to avoid possible joint model errors
df_ind$censor_time <- df_ind$censor_time + 0.0001
df_long$censor_time <- df_long$censor_time + 0.0001

# 3. Ensure data sets are ordered by ID then time for use in joint models
df_ind <- df_ind[order(df_ind$ID), ]
df_long <- df_long[order(df_long$ID, df_long$alsfrs_t_mnths), ]


# 4. Count number of longitudinal variables per individual
# Assign a variable to df2 to indicate how many measurements of each SNIP modality participants had
countlong <- data.frame(ID = names(table(df_long$ID)),
                    as.data.frame.matrix(matrix(table(df_long$ID))))
df_ind$countlong <- countlong[ match(df_ind$ID, countlong$ID), ]$V1


# 5. Calcuate adjusted timeline for this dataset - must adjust censor time and longtiudinal time
df_ind$adj_censortime <- df_ind$censor_time - df_ind$dx_delay
df_long$adj_censortime <- df_long$censor_time - df_long$dx_delay
df_long$adj_time <- df_long$alsfrs_t_mnths - df_long$dx_delay

# 6. Calculate time of first ALSFRS from diagnosis
df_ind$first_from_dx <- df_ind$First_alsfrs_t_mnths - df_ind$dx_delay


```


### Table 1. Demographics of study patients by diagnosis
```{r descriptives}
##### Descriptive Statistics
vars <- c("age", "dx_delay", "simp_site", "censor_time", "countlong",
          "First_alsfrs_t_mnths", "first_from_dx", "First_ALSFRS_Total")
factorVars <- c("simp_site")
tab1 <- CreateTableOne(vars = vars, factorVars = factorVars, data=df_ind,
                       test = TRUE, includeNA = TRUE)
tab1.export <- print(tab1, quote = FALSE, noSpaces = TRUE,
                     printToggle = FALSE, showAllLevels = TRUE,
                     nonnormal = c("dx_delay","censor_time", "First_alsfrs_t_mnths", "first_from_dx"))

print(tab1.export)
```


### Fit a lognormal distibution to delayed entry times for this cohort

```{r lognorm}
tidy ( fitdistr(df_ind$dx_delay, "lognormal") )

```
Note - these values differ slightly to those used in the simulation study because we were able to include more patients after we had run the simulation study.

### What is the mean value of the first ALSFRSR, mean time from onset, and mean time from diagnosis at time of first ALSFRSR
```{r firstalsfrs}
summary(df_ind$First_ALSFRS_Total)
summary(df_ind$First_alsfrs_t_mnths)
summary(df_ind$First_alsfrs_t_mnths - df_ind$dx_delay)

```

### Fit the Joint models used in the simulation study - A thru F
Note the for models E and F splines terms with 2 d.f. were used instead of polynomial terms.

```{r build_models, echo=FALSE, include = FALSE}
# Specify Cox models
cox_modA <- coxph (Surv (censor_time, censor) ~ simp_site + dx_delay,
                            data = df_ind, x = TRUE, model = TRUE)
cox_modB <- coxph (Surv (dx_delay, censor_time, censor) ~ simp_site + dx_delay,
                                 data = df_ind, x = TRUE, model = TRUE)
cox_modC <- coxph (Surv (adj_censortime, censor) ~ simp_site,
                                 data = df_ind, x = TRUE, model = TRUE)
cox_modD <- coxph (Surv (adj_censortime, censor) ~ simp_site + dx_delay,
                                 data = df_ind, x = TRUE, model = TRUE)
    
# Specify longitudinal model
lme_modA_B <- lme(Total.score ~ alsfrs_t_mnths,
                         random = ~ alsfrs_t_mnths | ID, data = df_long,
                         control = lmeControl(opt = "optim"))
lme_modC_D <- lme(Total.score ~ adj_time,
                         random = ~ adj_time | ID, data = df_long,
                         control = lmeControl(opt = "optim")) 

# Specify joint models
JM_A <- jointModelBayes(lme_modA_B, cox_modA, timeVar = "alsfrs_t_mnths")
JM_B <- jointModelBayes(lme_modA_B, cox_modB, timeVar = "alsfrs_t_mnths")
JM_C <- jointModelBayes(lme_modC_D, cox_modC, timeVar = "adj_time")
JM_D <- jointModelBayes(lme_modC_D, cox_modD, timeVar = "adj_time")

```
### Summary of Cox model hazard ratios
```{r cox_summ, echo= FALSE}
res_cox <- rbind( data.frame( Model = "A",
                       Parameter = tidy(cox_modA)$estimate,
                       tidy(cox_modA, exponentiate = TRUE) ),
                  data.frame( Model = "B",
                       Parameter = tidy(cox_modB)$estimate,
                       tidy(cox_modB, exponentiate = TRUE) ),
                  data.frame( Model = "C",
                       Parameter = tidy(cox_modC)$estimate,
                       tidy(cox_modC, exponentiate = TRUE) ),
                  data.frame( Model = "D",
                       Parameter = tidy(cox_modD)$estimate,
                       tidy(cox_modD, exponentiate = TRUE) ) )
res_cox <- res_cox %>% rename(  "Variable" = "term",
                                "HazRatio" = "estimate",
                               "CI_2.5" = "conf.low",
                               "CI_97.5" = "conf.high") %>% 
  arrange(Variable, Model)

kable(res_cox %>% dplyr::select(Variable, Model, Parameter, HazRatio, CI_2.5, CI_97.5, p.value), digits=3)


```

### Summary of JM event coefficients

```{r event_summary, echo= FALSE}
summ_JM_A <- summary(JM_A)
summ_JM_B <- summary(JM_B)
summ_JM_C <- summary(JM_C)
summ_JM_D <- summary(JM_D)


# Extract survival coefficients
event_coeffs <- rbind( data.frame(Model = "A", 
                                  Variable = row.names(summ_JM_A$`CoefTable-Event`),
                                  summ_JM_A$`CoefTable-Event`),
       data.frame(Model = "B", Variable = row.names(summ_JM_B$`CoefTable-Event`),
                  summ_JM_B$`CoefTable-Event`),
       data.frame(Model = "C", Variable = row.names(summ_JM_C$`CoefTable-Event`),
                  summ_JM_C$`CoefTable-Event`),
       data.frame(Model = "D", Variable = row.names(summ_JM_D$`CoefTable-Event`),
                  summ_JM_D$`CoefTable-Event`))
row.names(event_coeffs) <- NULL

# Drop Std Err and Std dev and convert value and CIs to exponential and arrange by variable first
event_coeffs <- event_coeffs %>%
  dplyr::select(-Std.Dev) %>% 
  mutate(HazRatio = exp(Value),
         CI_2.5 = exp(X2.5.),
         CI_97.5 = exp(X97.5.)) %>% 
  rename( "Parameter" = "Value") %>%
  filter(Variable != "tauBs") %>% 
  arrange(Variable, Model)

kable(event_coeffs %>% dplyr::select(Variable, Model, Parameter, Std.Err, HazRatio, CI_2.5, CI_97.5, P), digits=3)

```


### Summary of JM longtudinal coefficients
```{r long_summary, echo= FALSE}
# Extract longitudinal coefficients
event_coeffs_long <- rbind( data.frame(Model = "A", 
                                  Variable = row.names(summ_JM_A$`CoefTable-Long`),
                                  summ_JM_A$`CoefTable-Long`),
       data.frame(Model = "B", Variable = row.names(summ_JM_B$`CoefTable-Long`),
                  summ_JM_B$`CoefTable-Long`),
       data.frame(Model = "C", Variable = row.names(summ_JM_C$`CoefTable-Long`),
                  summ_JM_C$`CoefTable-Long`),
       data.frame(Model = "D", Variable = row.names(summ_JM_D$`CoefTable-Long`),
                  summ_JM_D$`CoefTable-Long`))
row.names(event_coeffs_long) <- NULL

# Drop Std Err and Std dev and convert value and CIs to exponential and arrange by variable first
event_coeffs_long <- event_coeffs_long %>%
  dplyr::select(-Std.Dev) %>% 
  rename( "Parameter" = "Value",
          "CI_2.5" = "X2.5.",
          "CI_97.5" = "X97.5.") %>%
  arrange(Variable, Model)

kable(event_coeffs_long %>% dplyr::select(Variable, Model, Parameter, Std.Err, CI_2.5, CI_97.5, P), digits=3)

```






